Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-18 paper2.qmd
section:5; process:write

<!-- recording: Sven and Tine 25.04.10; time: 00:16:30 - 00:19:00 -->

@@

<!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_scm], and the statistical model [-@fig-cj16_scm]. First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.

@@

Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16. We make this omission because, although these distributions contribute to the overall joint distribution of the data, they do not influence the relationship between $O_{R}$ and $D_{R}$. This independence stems from the random selection designs that govern $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$[^8].

@@

<!-- 2. Continuing with the assumptions -->
Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$. The discriminal processes represent the underlying written-quality trait of the texts. Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as: 
$$
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
$$ {#eq-mat52}
This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022]. In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases). It is important to clarify that the square brackets in $D_{R}$ merely indicate the relevant indices for each trait vector, without implying any subsetting of the data.

@@

Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$. We assume that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$. The parameter $p_{IA}$ is defined as a proportion of 1 to establish the scale of the latent trait, which is required in any latent variable model [@Depaoli_2021]. Note that because $e_{IA}$ follows a Normal distribution, $T_{IA}$ is also normally distributed. This result implies that our model reverts to assuming Normal discriminal processes for the stimuli, as originally proposed by @Thurstone_1927b (see @tbl-thurstone_cases).

@@

Fourth, we define $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$. As before, we assume that each element of $e_{I}$ is $iid$ Normal with a mean of zero. However, we consider the standard deviation, $s_{XI}$, varies depending on the teaching method group to which each student belongs. Given the example in @sec-theoretical, where the teaching method $X_{I} = 1,2$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$. Note that this parametrization of $e_{I}$ relaxes the equal dispersion assumption for the stimuli, effectively addressing the concerns raised in @sec-theory-issue1a.

@@

Fifth, we define $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments a judge makes affect the assessment of quality), and judgment-specific idiosyncratic errors $e_{JK}$. We assume that each element of $e_{JK}$ is $iid$ Normal with a mean of zero and a standard deviation $p_{JK}$, defined as a proportion of $1$, to anchor the scale of the latent trait [@Depaoli_2021].

Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., the effects of judgment expertise on the quality assessment) and judge-specific idiosyncratic errors $e_{J}$. Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs. For instance, if $Z_{J}$ is a variable that can take on three values representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].

@@

Finally, using *Bayesian inference methods* to transform the [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data. Bayesian inference offers three key advantages. First, it is well-suited for estimating complex and overparameterized models, where the number of parameters may exceed the number of observations [@Baker_1998; @Kim_et_al_1999]. Second, it allows researchers to incorporate prior information, which constrains parameter estimates within plausible bounds. This feature mitigates common estimation issues, such as non-convergence or improper solutions, often encountered in complex models under frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. Third, Bayesian inference is particularly effective for drawing conclusions from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014]. The **Declarations** section of this document provides a link to the model code, along with an alternative specification that assumes equal discriminal dispersions. We implemented both versions of the model using `Stan` [@Stan_2020, version 2.26.1]. Although the implementations were successful, we do not report the results here, as they fall outside the scope of this document.

@@

<!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:56:10 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->

<!-- Thus, this section addresses the issues identified in @sec-theory-issues by extending  Thurstone’s theory using the structural approach to causal inference. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis. -->

>>>>

<!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:16:30 - 00:19:00 -->

@@

<!-- 2. assumptions about the outcome -->
Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_scm], and the statistical model [-@fig-cj16_scm]. First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.

@@

<!-- 3. Note about the sample and comparison mechanism, as well as the predictor's distribution -->
Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16. We make this omission because, although these distributions contribute to the overall joint distribution of the data, they do not influence the relationship between $O_{R}$ and $D_{R}$. This independence stems from the random selection designs that govern $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$[^8].

@@

<!-- 4. functional form for D_{R} -->
Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$. The discriminal processes represent the underlying written-quality trait of the texts. Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as: 
$$
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
$$ {#eq-mat52}
This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022]. In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases). It is important to clarify that the square brackets in $D_{R}$ merely indicate the relevant indices for each trait vector, without implying any subsetting of the data.

@@

<!-- 5. Functional form and assumptions for T_{IA} -->
Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$. We assume that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$. The parameter $p_{IA}$ is defined as a proportion of 1 to establish the scale of the latent trait, which is required in any latent variable model [@Depaoli_2021]. Note that because $e_{IA}$ follows a Normal distribution, $T_{IA}$ is also normally distributed. This result implies that our model reverts to assuming Normal discriminal processes for the stimuli, as originally proposed by @Thurstone_1927b (see @tbl-thurstone_cases).

@@

<!-- 6. Functional form and assumptions for T_{I} -->
Fourth, we define $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$. As before, we assume that each element of $e_{I}$ is $iid$ Normal with a mean of zero. However, we consider the standard deviation, $s_{XI}$, varies depending on the teaching method group to which each student belongs. Given the example in @sec-theoretical, where the teaching method $X_{I} = 1,2$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$. Note that this parametrization of $e_{I}$ relaxes the equal dispersion assumption for the stimuli, effectively addressing the concerns raised in @sec-theory-issue1a.

@@

<!-- 7. Functional form and assumptions for B_{JK} -->
Fifth, we define $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments a judge makes affect the assessment of quality), and judgment-specific idiosyncratic errors $e_{JK}$. We assume that each element of $e_{JK}$ is $iid$ Normal with a mean of zero and a standard deviation $p_{JK}$, defined as a proportion of $1$, to anchor the scale of the latent trait [@Depaoli_2021].

<!-- 8. Functional form and assumptions for B_{J} -->
Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., the effects of judgment expertise on the quality assessment) and judge-specific idiosyncratic errors $e_{J}$. Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs. For instance, if $Z_{J}$ is a variable that can take on three values representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].

@@

<!-- 9. Computational implementation -->
Finally, we use *Bayesian inference methods* to convert [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data. Bayesian inference offers three key advantages for this purpose. First, it handles complex and overparameterized models--such as the one we implement--where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999]. Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. In our implementation, we use priors to define the error distribution, which sets the scale of the latent variables [@Depaoli_2014]. Third, Bayesian inference supports robust inferences from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014]. This feature is particularly relevant in CJ experiments, where researchers often collect large volumes of paired comparisons but work with relatively small samples of judges, stimuli, and individuals to test hypotheses.

The **Declarations** section of this document provides a link to the model code, along with an alternative specification that assumes equal discriminal dispersions. We implemented both versions of the model using `Stan` [@Stan_2020, version 2.26.1]. Although the implementations were successful, we do not report the results here, as they fall outside the scope of this document.

@@

<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:58:20 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->

<!-- make a summary of what you have accomplished so far -->

<!-- change the paradigm -->

<!-- opportunities open  -->

<!-- Now that we have this model we have a principled way to make some power analysis which will helps us to design studies, rather than just relying in simple rules of thumbs. -->

<!-- Thus, this study addresses the issues identified in @sec-theory-issues by extending  Thurstone’s general form using the structural approach to causal inference and Bayesian inference methods. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons, i.e, the sampleand compariosn mechanisms. In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis. -->
